---
title: "RAP exemplar"
output:
  html_document: default
  html_notebook: default
---

## Reproducible Analytical Pipeline

Producing official statistics for publications is a key function of many teams across government. Here we showcase the exemplar `eesectors` package which can be installed to your machine from Github.  

For why using a reproducible analytical pipeline is useful, see this Government Digital Service (GDS) [blog post](https://gdsdata.blog.gov.uk/2017/03/27/reproducible-analytical-pipeline/).  

### Installation

The package can then be installed using
`devtools::install_github('ukgovdatascience/eesectors')`. Some users may
not be able to use the `devtools::install_github()` commands as a result
of network security settings. If this is the case, `eesectors` can be
installed by downloading the [zip of the
repository](https://github.com/ukgovdatascience/govstyle/archive/master.zip)
and installing the package locally using
`devtools::install_local(<path to zip file>)`.

#### Version control

As the code is stored on Github I can access the current master version as well as all [historic versions](https://github.com/ukgovdatascience/eesectors/releases). This allows me to reproduce a report from last year for example. I can look at what release version was use and install that accordingly using the [additional arguments](ftp://cran.r-project.org/pub/R/web/packages/githubinstall/vignettes/githubinstall.html) for `install_github`. 

### Loading the package

Installation means the package is on our computer but it is not loaded into the computer's working memory. We also load any additional packages that might be useful for exploring the package or data therein.    

```{r}

library(eesectors)
library(tidyverse)

```

This makes all the functions within the package available for use. It also provides us with some R [data objects](https://github.com/ukgovdatascience/eesectors/tree/master/data), such as aggregated data sets ready for visualisations or analysis within the report.  

> Packages are the fundamental units of reproducible R code. They include reusable R functions, the documentation that describes how to use them, and sample data. - Hadely Wickham

### Explore the package

A good place to start is the package [README](https://github.com/ukgovdatascience/eesectors).

#### Status badges

The [status badges](https://stackoverflow.com/questions/35563012/what-are-the-status-tags-like-build-passing) provide useful information. They are found in the top left of the README and should be green and say passing. This indicates that this package will run OK on Windows and linux or mac. Essentially the package is likely to build correctly on your machine when you install it.  

More detail concerning the softeware development angle can be found in Matt Upson's blog post.

#### Look at the output first

If you go to chapter 3 it is apparent that most of the content is either data tables of summary statistics or visualisation of the data. This makes automation particularly useful here and likely to make time savings. Chapter 3 seems to be fairly typical in its length (if not a bit shroter compared to other chapters).

This package seems to work by taking the necessary data inputs as arguments in a function then outputting the relevant figures. The names of the functions match the figures they produce. Prior to this step we have to get the data in the correct format

The `GVA_by_sector_2016` provides test data to work with during development. This will be important for the development of other packages for different reports. You need a precise understanding of how you go from raw data, to aggregated data (such as `GVA_by_sector_2016`) to the final figure. What are your inputs (arguments) and outputs?

```{r}
glimpse(GVA_by_sector_2016)
x <- GVA_by_sector_2016
```

#### Automating QA

Human's are not particularly good at QA, especially when working with massive spreadsheets it's easy for errors to creep in. We can automate alot of the sense checking and update this if things change or a human provides another creative test to use for sense checking. If you can describe the test to a colleague then you can code it.  

The author uses messages to tell us what checks are being conducted or we can look at the body of the function if we are interested. This is useful if you are considering developing your own package, it will help you struture the message which are useful for the user.

```{r}

gva <- year_sector_data(GVA_by_sector_2016)
```

This is a semi-automated process so the user should check the Checks and ensure they meet their usual checks that would be conducted manually. If a new check or test becomes necessary then it should be implemented by changing the code.

```{r}
body(year_sector_data)
```

The function is  structured to tell the user what check is being made and then running that check given the input `x`. If the input fails a check the function is stopped with a useful diagnostic message for the user. This is achieved using `if` and the opposite of the desired feature of `x`.  

```{r eval=FALSE}
    message("Checking x has correct columns...")
    if (length(colnames(x)) != 3) 
        stop("x must have three columns: sector, year, and one of GVA, export, or x")
```


For example, if `x` does not have exactly three columns we `stop`.

#### output of this function

The output object is different to the input as expected, yet it does contain the initial data.

```{r}
identical(gva$df, x)

```

The rest of the list contains other details that could be changed at a later date if required, demonstrating defensive programming. For example, the sectors that are of interest to DCMS have changed and may change again.

```{r eval=FALSE}
?year_sector_data
```


Let's take a closer look at this function using the help and other standard function exploration functions.

The help says it produces a custom class of object with five slots.

```{r}
isS4(gva)
class(gva)
```

It's not actually an S4 object, by slots the author means a list of objects.

#### The input

The input, which is likely a bunch of [not tidy or messy](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) spreadsheets needs to be wrangled and aggregated (if necessary) for input into the functions prefixed by `figure`.

```{r}
glimpse(GVA_by_sector_2016)
```

#### The R output

With the data in the appropraite form to be received as an argument or input for the `figure` family of functions we can proceed to plot.

```{r}
figure3.1(x = gva)

```

Again we can look at the details of the plot. We could change the body of the function to affect change to the default plot or we can pass additional `ggplot` arguments to it.

Reading the code we see it filters the data, makes the variables it needs, refactors the `sector` variable and then plots it. 

```{r}
body(figure3.1)

```

I can inspect and change an argument if I feel inclined or if a new colour scheme becomes preferred for example. However, there is no `...` in the body of the function itself so where does this argument get passed to?

This all looks straight forward and we can inspect the other functions for generating the figures or plot output.

```{r}
body(figure3.2)
body(figure3.3)
```

#### Error handling

A point of interest in the code with which some users may be unfamiliar is `tryCatch` which is a function that allows the function to catch conditions such as warnings, errors and messages. We see this towards the end of the function where if either of these conditions are produced then an informative message is produced (in that it tells you in what function there was a problem). The structure here is simple and could be copied and pasted for use in automating other figures of other chapters or statistical reports.  

For a comprehensive introduction see [Hadley's Chapter](http://adv-r.had.co.nz/Exceptions-Debugging.html#condition-handling).  

## Take home message

We have explored the `eesectors` package from the perspective of someone wishing to develop our own semi-automated chapter production through the development of a package in R. This package provides a useful tempplate where one could copy the foundations of the package and workflow.  

```{r}
sessionInfo()
```


